Climate variability indices for ecological and crop models in R: the \texttt{climatrends} package

### 2020-05-26 inst/paper/paper.Rmd
bean_plot

---
title: 'Climate variability indices for ecological and crop models in R: the `climatrends` package'
tags:
- climate data
- climatology
- earth science
- evapotranspiration
- precipitation data
- R
- reproducibility
- weather data
authors:
  - name: Kauê de Sousa
    orcid: 0000-0002-7571-7845
    affiliation: "1, 2"
  - name: Jacob van Etten
    orcid: 0000-0001-7554-2558
    affiliation: 2
  - name: Svein Ø. Solberg
    orcid: 0000-0002-4491-4483
    affiliation: 1
affiliations:
  - name: Department of Agricultural Sciences, Inland Norway University of Applied Sciences, Hamar, Norway
    index: 1
  - name: Bioversity International, Rome, Italy
    index: 2
citation_author: de Sousa et. al.
date: "`r format(Sys.time(), '%d %B %Y')`"
year: 2020
bibliography: paper.bib
output:
  pdf_document:
    keep_tex: true
header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

@@

Abiotic factors plays an important role in most ecological and crop systems that depends on certain levels of temperature, light and precipitation (and their interplay) to initiate important physiological events [@PlantEcology]. In the walk of climate change, understand how these factors drives the physiological processes is a key approach to provide recommendations for adaptation and biodiversity conservation. If translated to stress

Raw temperature and precipitation values are ... to describe how organisms interacts with the enviroment

@@

@@

For tropical areas, where the temperature may surpass a maximum threshold ($T_{0^{max}}$), resulting in limited development, the minimum temperature is adjusted using Equation 3 and the maximum temperature is adjusted to a maximum base temperature as follow.

@@

$$ T_{max} > T_{0^{max}} \; \rightarrow \; T_{max} = T_{0^{max}} $$

Where $T_{0^{max}}$ is the maximum base temperature for growth, defined in `GDD()` using the argument `tbase_max`.

@@

@@

Two functions in **climatrends** are mainly designed to capture the effects of climate on the development and stress of crop species, `crop_sensitive` computes indices that aims to capture the changes in temperature extremes during key phenological stages (e.g. anthesis), and `ETo()` computes the reference evapotranspiration.

The reference evapotranspiration measures the influence of the climate on a given organism water needs, generally a crop species [@Brouwer1986]. The function `ETo()` applies the Blaney-Criddle method, a general theoretical method used when only air-temperature is available locally. It should be noted that this method is not very accurate and aims to provide the order of magnitude of evapotranspitation. The reference evapotranspiration is calculated using the following equation.

@@

## Seed germination or some GDD related analysis

Use the data from seed germination or crop growth to compute GDD. How many GDD a seed need to become a seedling?

@@

During five growing seasons (from 2015 to 2016 in Nicaragua), van Etten et al [-@vanEtten2019] conducted a crowdsourcing citizen-science trial testing 11 common beans varieties (*Phaseolus vulgaris* L.) in 842 farmer-managed experimental plots. Sets of three varieties were allocated randomly to farms as incomplete blocks, maintaining spatial balance by assigning roughly equal frequencies of the varieties to each area. For the analysis the authors applied the Plackett–Luce model, which estimates for each variety the probability that it wins, beating all other varieties in the set [@Turner2020]. The research showed that this approach can register known specific effects of climate variation on crop varietal performance. An earlier version of `climatrends` was used in this research to capture the seasonal climate variation, here we replicate part of this analysis in the matter that concerns the calculation and application of the climate indices.

Is important to mention that this analysis shaped the implementation of the methods `temperature.array()` and `rainfall.matrix()`.

@@

@@

# fit a Plackett-Luce tree
plt <- pltree(G ~ maxNT, data = cbean, minsize = 200)
```

@@

```{r fig_cbean, fig.cap="Fig. 1. Application of climatrends functions to support the analysis of a citizen-science data testing 11 common bean varieties in Nicaragua. (A) Days required to reach 900 growing-degree days from planting date calculated using the function GDD(). (B) Maximum night temperature (°C) distributed across seasons computed using the function temperature(). (C) Plackett-Luce Tree showing the probability of one common bean variety has to win against the others (axys X) in two different nodes splitted with the maximum night temperature (°C).", out.width = '80%', echo=FALSE}
knitr::include_graphics("cbean.png")
```

>>>>

---
title: 'Climate variability indices for ecological and crop models in R: the `climatrends` package'
tags:
- climate data
- climatology
- earth science
- evapotranspiration
- precipitation data
- R
- reproducibility
- weather data
authors:
  - name: Kauê de Sousa
    orcid: 0000-0002-7571-7845
    affiliation: "1, 2"
  - name: Jacob van Etten
    orcid: 0000-0001-7554-2558
    affiliation: 2
  - name: Svein Ø. Solberg
    orcid: 0000-0002-4491-4483
    affiliation: 1
affiliations:
  - name: Department of Agricultural Sciences, Inland Norway University of Applied Sciences, Hamar, Norway
    index: 1
  - name: Bioversity International, Rome, Italy
    index: 2
citation_author: de Sousa et. al.
date: "`r format(Sys.time(), '%d %B %Y')`"
year: 2020
bibliography: paper.bib
csl: citationstyle.csl
output:
  pdf_document:
    keep_tex: true
header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

@@

Abiotic factors plays an important role in most ecological and crop systems that depends on certain levels of temperature, light and precipitation (and their interplay) to initiate important physiological events [@PlantEcology]. In the walk of climate change, understand how these factors drives the physiological processes is a key approach to provide recommendations for adaptation and biodiversity conservation.

@@

Some explanation on the `temperature()` and `rainfall()` functions.

@@

For tropical areas, where the temperature may surpass a maximum threshold ($T_{0_{max}}$), resulting in limited development, the minimum temperature is adjusted using Equation 3 and the maximum temperature is adjusted to a maximum base temperature as follow.

@@

$$ T_{max} > T_{0_{max}} \; \rightarrow \; T_{max} = T_{0_{max}} $$

Where $T_{0_{max}}$ is the maximum base temperature for growth, defined in `GDD()` using the argument `tbase_max`.

@@

Explain how the `late_frost()` is implemented.

@@

Two functions in **climatrends** are mainly designed to capture the effects of climate on the development and stress of crop species, `crop_sensitive()` computes indices that aims to capture the changes in temperature extremes during key phenological stages (e.g. anthesis), and `ETo()` computes the reference evapotranspiration.

The reference evapotranspiration measures the influence of the climate on a given plant water needs [@Brouwer1986]. The function `ETo()` applies the Blaney-Criddle method, a general theoretical method used when only air-temperature is available locally. It should be noted that this method is not very accurate and aims to provide the order of magnitude of evapotranspitation. The reference evapotranspiration is calculated using the following equation.

@@

## Some GDD related analysis

Use the data from seed germination or crop growth to compute GDD. Maybe some GDD analysis in Scandinavia with data from the reindeer project.

@@

During five growing seasons (from 2015 to 2017 in Nicaragua), van Etten et al. [-@vanEtten2019] conducted a crowdsourcing citizen-science trial testing 11 common beans varieties (*Phaseolus vulgaris* L.) in 842 farmer-managed experimental plots. Sets of three varieties were allocated randomly to farms as incomplete blocks. A Plackett–Luce model was used to analyse the data, this model estimates for each variety the probability that it wins, beating all other varieties in the set [@Turner2020]. An earlier version of `climatrends` was used in this research to capture the seasonal climate variation, here we reproduce part of this analysis regarding calculation and application of the climate indices. The approach here is slightly different because if considers the growing-degree days from planting date to maturity (the earlier study used planting date to the end of reproductive stage) and add new indices to illustrate the package implementation.

The data used here is available as supplementary materials (**zenodo ref**) as `cbean`. This contains a data.frame with a Plackett-Luce grouped rankings, the geographical coordinates of each sampled plot and the planting dates from where each farmer decided to start the experiment. The dates can differ from each other in the same season. The temperature data used was the land surface temperature MODIS (MYD11A2) [@Wan2015] and is storaged as an array with two layers (1st for the day and 2nd for the night temperatures). Each column corresponds to the dates (from 2015-09-10 to 2017-06-09) and the rows corresponds to the rows in the `cbean` data.frame.

Since the phenological stages were not available, we estimate these stages based on the amount of growing degree-days required to reach a given stage using the function `GDD()`. For common beans, we define 900 degree-days, from planting date to maturity [@deMedeiros2016]. The input data is the array with the temperature data, the vector with planting dates (`cbean$planting_date`), the required amount of degree-days passed to the argument `degree.days` and the character string 'ndays' specifying that the function must return the values as number of days. `GDD()` calls internally the function `get_timeseries()` which will match the given dates in `day.one` with the column names in the array and concatenate the values for each row. Then `GDD()` computes the degree-days for the time series and return the length of the vector where the accumulated gdd reached the pre-defined threshold (900).

The degree-days spanned from 54 to 100 days as shown in Fig. 2a. For simplicity we take the average per season and use this vector to compute the temperature indices.

@@

To compute the temperature indices we use the array with tempeture data, the vector with planting dates, and the seasonal averaged degree-days passed as a vector using the argument `span`. `temperature()` concatenates the data from the given `day.one` to the given `span` and compute the indices for each row.

In van Etten [-@vanEtten2019], a forward variable selection was applied to retain the most representative covariates based on the deviance reduction. This analysis retained the maximum night temperature (maxNT) as the most representative covariate. To illustrate how the Plackett-Luce trees can grow in complexity as we add more indices, we included the summer days (SU, number of days with maximum day temperature > 30 $^\circ C$) together with maxNT.

@@

# fit a Plackett-Luce tree
plt <- pltree(G ~ maxNT + SU, data = cbean, minsize = 50)
```

Across-season distribution of maxNT captured for each sample plot in this experiment is shown in Fig. 2b. The data has a bimodal distribution which is reflected in the splitting value (18.7 $^\circ C$) for the Plackett-Luce trees in Fig. 2c. The upper node splits with 49 summer days (SU). We can interpret these results as that differences in growing performance of common beans is led by a considerable amount of diurnal temperature above a warmer threshold of 30 $^\circ C$ (in this case >70% of the growing days) and warmer nights (> 18.7 $^\circ C$).

@@

```{r fig_cbean, fig.cap="Fig. 1. Application of climatrends functions to support the analysis of a citizen-science data testing 11 common bean varieties in Nicaragua. (A) Days required to reach 900 growing-degree days from planting date calculated using the function GDD(). (B) Maximum night temperature (°C) distributed across seasons computed using the function temperature(). (C) Plackett-Luce Tree showing the probability of one common bean variety has to win against the others (axys X) in three different nodes splitted with the summer days (day temperature > 30 °C) and maximum night temperature (°C). Note: the first season (primera, Pr) spans from May to August, the second (postrera, Po) from September to October, and the third (apante, Ap) from November to January.", out.width = '80%', echo=FALSE}
knitr::include_graphics("cbean.png")
```

### 2020-05-26 inst/paper/paper.tex
bean_plot

Abiotic factors plays an important role in most ecological and crop
systems that depends on certain levels of temperature, light and
precipitation (and their interplay) to initiate important physiological
events (Schulze et al. 2019). In the walk of climate change, understand
how these factors drives the physiological processes is a key approach
to provide recommendations for adaptation and biodiversity conservation.
If translated to stress

Raw temperature and precipitation values are \ldots{} to describe how
organisms interacts with the enviroment

\texttt{climatrends} aims to provide the R (R Core Team 2020) toolkit to
compute extreme precipitation and temperature indices that serves as
input for climate and crop models (van Etten et al. 2019; Kehel, Crossa,
and Reynolds 2016), trends in climate change (Aguilar et al. 2005; de
Sousa et al. 2018) and applied ecology (Prentice et al. 1992; Liu and
El-Kassaby 2018).

@@

Six main functions are provided, \texttt{crop\_sensitive()},
\texttt{ETo()}, \texttt{GDD()}, \texttt{late\_frost()},
\texttt{rainfall()} and \texttt{temperature()} with a default method for
numeric `vector's and additional methods implemented via the package
\texttt{methods} (R Core Team 2020) for classes 'matrix' (or array),
`data.frame', and `sf' (of geometry POINT or POLYGON) (Pebesma 2018).
The later two are designed to fetch data from cloud sources.

@@

@@

Growing degree-days (gdd) is an heuristic tool in phenology that
measures heat accumulation and is used to predict plant and animal
development rates (Prentice et al. 1992). Growing degree-days are
calculated by taking the integral of warmth above a base temperature
(\(T_{0}\)). The function \texttt{GDD()} applies by default the
following equation.

@@

By default, the function returns the degree-days that is accumulated
over the time series using Equation 1. Additionally, the function may
return the daily values of degree-days or the number of days that a
given organism required to reach a certain number of accumulated
degree-days. These values are defined by `acc', `daily' or `ndays' and
can be adjusted using the argument \texttt{return.as}. The required
accumulated gdd is defined with argument \texttt{degree.days}. For
example, the Korean pine (\emph{Pinus koraiensis}) requires 105
\(^\circ C\) accumulated gdd to onset the photosynthesis (Wu et al.
2013). In that case, \texttt{GDD()} will calculate the growing
degree-days (\(gdd\)) and sum up the values until it reaches 105
\(^\circ C\) and return the number of days required in the given season
(\(GDD_{r}\)), as follows.

@@

@@

The reference evapotranspiration measures the influence of the climate
on a given organism water needs, generally a crop species (Brouwer and
Heibloem 1986). The function \texttt{ETo()} applies the Blaney-Criddle
method, a general theoretical method used when only air-temperature is
available locally. It should be noted that this method is not very
accurate and aims to provide the order of magnitude of
evapotranspitation. The reference evapotranspiration is calculated using
the following equation.

@@

\hypertarget{seed-germination-or-some-gdd-related-analysis}{%
\subsection{Seed germination or some GDD related
analysis}\label{seed-germination-or-some-gdd-related-analysis}}

Use the data from seed germination or crop growth to compute GDD. How
many GDD a seed need to become a seedling?

@@

During five growing seasons (from 2015 to 2016 in Nicaragua), van Etten
et al (2019) conducted a crowdsourcing citizen-science trial testing 11
common beans varieties (\emph{Phaseolus vulgaris} L.) in 842
farmer-managed experimental plots. Sets of three varieties were
allocated randomly to farms as incomplete blocks, maintaining spatial
balance by assigning roughly equal frequencies of the varieties to each
area. For the analysis the authors applied the Plackett--Luce model,
which estimates for each variety the probability that it wins, beating
all other varieties in the set (Turner et al. 2020). The research showed
that this approach can register known specific effects of climate
variation on crop varietal performance. An earlier version of
\texttt{climatrends} was used in this research to capture the seasonal
climate variation, here we replicate part of this analysis in the matter
that concerns the calculation and application of the climate indices.

Is important to mention that this analysis shaped the implementation of
the methods \texttt{temperature.array()} and \texttt{rainfall.matrix()}.

@@

@@

\CommentTok{\# fit a Plackett{-}Luce tree}
\NormalTok{plt <{-}}\StringTok{ }\KeywordTok{pltree}\NormalTok{(G }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{maxNT, }\DataTypeTok{data =}\NormalTok{ cbean, }\DataTypeTok{minsize =} \DecValTok{200}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-Aguilar2005}{}%
Aguilar, E., T. C. Peterson, P. Ramírez Obando, R. Frutos, J. A. Retana,
M. Solera, J. Soley, et al. 2005. ``Changes in precipitation and
temperature extremes in Central America and northern South America,
1961--2003.'' \emph{Journal of Geophysical Research} 110 (D23): D23107.
\url{https://doi.org/10.1029/2005JD006119}.

\leavevmode\hypertarget{ref-Brouwer1986}{}%
Brouwer, C., and M. Heibloem. 1986. \emph{Irrigation water management:
Irrigation water needs}. Training m. Rome, Italy: Food; Agriculture
Organization of The United Nations.
\url{http://www.fao.org/3/S2022E/s2022e00.htm}.

\leavevmode\hypertarget{ref-deSousa2018}{}%
de Sousa, Kauê, Fernando Casanoves, Jorge Sellare, Alejandra Ospina,
Jose Gabriel Suchini, Amilcar Aguilar, and Leida Mercado. 2018. ``How
climate awareness influences farmers' adaptation decisions in Central
America?'' \emph{Journal of Rural Studies} 64: 11--19.
\url{https://doi.org/10.1016/j.jrurstud.2018.09.018}.

\leavevmode\hypertarget{ref-Kehel2016}{}%
Kehel, Z., J. Crossa, and M. Reynolds. 2016. ``Identifying Climate
Patterns during the Crop-Growing Cycle from 30 Years of CIMMYT Elite
Spring Wheat International Yield Trials.'' In \emph{Applied Mathematics
and Omics to Assess Crop Genetic Resources for Climate Change Adaptive
Traits}, edited by Abdallah Bari, Ardeshir B. Damania, Michael Mackay,
and Selvadurai Dayanandan, 151--74. CRC Press.

\leavevmode\hypertarget{ref-YLiu2018}{}%
Liu, Yang, and Yousry A. El-Kassaby. 2018. ``Evapotranspiration and
favorable growing degree-days are key to tree height growth and
ecosystem functioning: Meta-analyses of Pacific Northwest historical
data.'' \emph{Scientific Reports} 8228 (October).
\url{https://doi.org/10.1038/s41598-018-26681-1}.

\leavevmode\hypertarget{ref-sf}{}%
Pebesma, Edzer. 2018. ``Simple Features for R: Standardized Support for
Spatial Vector Data.'' \emph{The R Journal} 10 (1): 439--46.
\url{https://doi.org/10.32614/RJ-2018-009}.

\leavevmode\hypertarget{ref-Prentice1992}{}%
Prentice, I. Colin, Wolfgang Cramer, Sandy P. Harrison, Rik Leemans,
Robert A. Monserud, and Allen M. Solomon. 1992. ``Special Paper: A
Global Biome Model Based on Plant Physiology and Dominance, Soil
Properties and Climate.'' \emph{Journal of Biogeography} 19 (2): 117.
\url{https://doi.org/10.2307/2845499}.

\leavevmode\hypertarget{ref-RCoreTeam}{}%
R Core Team. 2020. ``R: A language and environment for statistical
computing. version 4.0.0.'' Vienna, Austria: CRAN R Project.
\url{https://www.r-project.org/}.

\leavevmode\hypertarget{ref-PlantEcology}{}%
Schulze, Ernst-Detlef, Erwin Beck, Nina Buchmann, Stephan Clemens, Klaus
Müller-Hohenstein, and Michael Scherer-Lorenzen. 2019. \emph{Plant
Ecology}. Second Edi. Berlin, Germany: Springer.
\url{https://doi.org/10.1007/978-3-662-56233-8}.

\leavevmode\hypertarget{ref-Turner2020}{}%
Turner, Heather L, Jacob van Etten, David Firth, and Ioannis Kosmidis.
2020. ``Modelling rankings in R: the PlackettLuce package.''
\emph{Computational Statistics}.
\url{https://doi.org/10.1007/s00180-020-00959-3}.

\leavevmode\hypertarget{ref-vanEtten2019}{}%
van Etten, Jacob, Kauê de Sousa, Amílcar Aguilar, Mirna Barrios, Allan
Coto, Matteo Dell'Acqua, Carlo Fadda, et al. 2019. ``Crop variety
management for climate adaptation supported by citizen science.''
\emph{Proceedings of the National Academy of Sciences} 116 (10):
4194--9. \url{https://doi.org/10.1073/pnas.1813720116}.

\leavevmode\hypertarget{ref-JWu2013}{}%
Wu, Jiabing, Dexin Guan, Fenhui Yuan, Anzhi Wang, and Changjie Jin.
2013. ``Soil Temperature Triggers the Onset of Photosynthesis in Korean
Pine.'' Edited by Carl J. Bernacchi. \emph{PLoS ONE} 8 (6): e65401.
\url{https://doi.org/10.1371/journal.pone.0065401}.
\end{cslreferences}

@@

\begin{figure}
\includegraphics[width=0.8\linewidth]{cbean} \caption{Fig. 1. Application of climatrends functions to support the analysis of a citizen-science data testing 11 common bean varieties in Nicaragua. (A) Days required to reach 900 growing-degree days from planting date calculated using the function GDD(). (B) Maximum night temperature (°C) distributed across seasons computed using the function temperature(). (C) Plackett-Luce Tree showing the probability of one common bean variety has to win against the others (axys X) in two different nodes splitted with the maximum night temperature (°C).}\label{fig:fig_cbean}
\end{figure}

>>>>

Abiotic factors plays an important role in most ecological and crop
systems that depends on certain levels of temperature, light and
precipitation (and their interplay) to initiate important physiological
events (Schulze et al., 2019). In the walk of climate change, understand
how these factors drives the physiological processes is a key approach
to provide recommendations for adaptation and biodiversity conservation.

\texttt{climatrends} aims to provide the R (R Core Team, 2020) toolkit
to compute extreme precipitation and temperature indices that serves as
input for climate and crop models (Kehel, Crossa, \& Reynolds, 2016; van
Etten et al., 2019), trends in climate change (Aguilar et al., 2005; de
Sousa et al., 2018) and applied ecology (Prentice et al., 1992; Liu \&
El-Kassaby, 2018).

@@

Six main functions are provided, \texttt{crop\_sensitive()},
\texttt{ETo()}, \texttt{GDD()}, \texttt{late\_frost()},
\texttt{rainfall()} and \texttt{temperature()} with a default method for
numeric `vector's and additional methods implemented via the package
\texttt{methods} (R Core Team, 2020) for classes 'matrix' (or array),
`data.frame', and `sf' (of geometry POINT or POLYGON) (Pebesma, 2018).
The later two are designed to fetch data from cloud sources.

@@

Some explanation on the \texttt{temperature()} and \texttt{rainfall()}
functions.

@@

Growing degree-days (gdd) is an heuristic tool in phenology that
measures heat accumulation and is used to predict plant and animal
development rates (Prentice et al., 1992). Growing degree-days are
calculated by taking the integral of warmth above a base temperature
(\(T_{0}\)). The function \texttt{GDD()} applies by default the
following equation.

@@

By default, the function returns the degree-days that is accumulated
over the time series using Equation 1. Additionally, the function may
return the daily values of degree-days or the number of days that a
given organism required to reach a certain number of accumulated
degree-days. These values are defined by `acc', `daily' or `ndays' and
can be adjusted using the argument \texttt{return.as}. The required
accumulated gdd is defined with argument \texttt{degree.days}. For
example, the Korean pine (\emph{Pinus koraiensis}) requires 105
\(^\circ C\) accumulated gdd to onset the photosynthesis (Wu, Guan,
Yuan, Wang, \& Jin, 2013). In that case, \texttt{GDD()} will calculate
the growing degree-days (\(gdd\)) and sum up the values until it reaches
105 \(^\circ C\) and return the number of days required in the given
season (\(GDD_{r}\)), as follows.

@@

Explain how the \texttt{late\_frost()} is implemented.

@@

The reference evapotranspiration measures the influence of the climate
on a given organism water needs, generally a crop species (Brouwer \&
Heibloem, 1986). The function \texttt{ETo()} applies the Blaney-Criddle
method, a general theoretical method used when only air-temperature is
available locally. It should be noted that this method is not very
accurate and aims to provide the order of magnitude of
evapotranspitation. The reference evapotranspiration is calculated using
the following equation.

@@

\hypertarget{some-gdd-related-analysis}{%
\subsection{Some GDD related analysis}\label{some-gdd-related-analysis}}

Use the data from seed germination or crop growth to compute GDD. Maybe
some GDD analysis in Scandinavia with data from the reindeer project.

@@

During five growing seasons (from 2015 to 2017 in Nicaragua), van Etten
et al.~(2019) conducted a crowdsourcing citizen-science trial testing 11
common beans varieties (\emph{Phaseolus vulgaris} L.) in 842
farmer-managed experimental plots. Sets of three varieties were
allocated randomly to farms as incomplete blocks. A Plackett--Luce model
was used to analyse the data, this model estimates for each variety the
probability that it wins, beating all other varieties in the set
(Turner, van Etten, Firth, \& Kosmidis, 2020). An earlier version of
\texttt{climatrends} was used in this research to capture the seasonal
climate variation, here we reproduce part of this analysis regarding
calculation and application of the climate indices. The approach here is
slightly different because if considers the growing-degree days from
planting date to maturity (the earlier study used planting date to the
end of reproductive stage) and add new indices to illustrate the package
implementation.

The data used here is available as supplementary materials
(\textbf{zenodo ref}) as \texttt{cbean}. This contains a data.frame with
a Plackett-Luce grouped rankings, the geographical coordinates of each
sampled plot and the planting dates from where each farmer decided to
start the experiment. The dates can differ from each other in the same
season. The temperature data used was the land surface temperature MODIS
(MYD11A2) (Wan, Hook, \& Hulley, 2015) and is storaged as an array with
two layers (1st for the day and 2nd for the night temperatures). Each
column corresponds to the dates (from 2015-09-10 to 2017-06-09) and the
rows corresponds to the rows in the \texttt{cbean} data.frame.

Since the phenological stages were not available, we estimate these
stages based on the amount of growing degree-days required to reach a
given stage using the function \texttt{GDD()}. For common beans, we
define 900 degree-days, from planting date to maturity (de Medeiros,
Daniel, \& Fengler, 2016). The input data is the array with the
temperature data, the vector with planting dates
(\texttt{cbean\$planting\_date}), the required amount of degree-days
passed to the argument \texttt{degree.days} and the character string
`ndays' specifying that the function must return the values as number of
days. \texttt{GDD()} calls internally the function
\texttt{get\_timeseries()} which will match the given dates in
\texttt{day.one} with the column names in the array and concatenate the
values for each row. Then \texttt{GDD()} computes the degree-days for
the time series and return the length of the vector where the
accumulated gdd reached the pre-defined threshold (900).

The degree-days spanned from 54 to 100 days as shown in Fig. 2a. For
simplicity we take the average per season and use this vector to compute
the temperature indices.

@@

To compute the temperature indices we use the array with tempeture data,
the vector with planting dates, and the seasonal averaged degree-days
passed as a vector using the argument \texttt{span}.
\texttt{temperature()} concatenates the data from the given
\texttt{day.one} to the given \texttt{span} and compute the indices for
each row.

In van Etten (2019), a forward variable selection was applied to retain
the most representative covariates based on the deviance reduction. This
analysis retained the maximum night temperature (maxNT) as the most
representative covariate. To illustrate how the Plackett-Luce trees can
grow in complexity as we add more indices, we included the summer days
(SU, number of days with maximum day temperature \textgreater{} 30
\(^\circ C\)) together with maxNT.

@@

\CommentTok{\# fit a Plackett{-}Luce tree}
\NormalTok{plt <{-}}\StringTok{ }\KeywordTok{pltree}\NormalTok{(G }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{maxNT }\OperatorTok{+}\StringTok{ }\NormalTok{SU, }\DataTypeTok{data =}\NormalTok{ cbean, }\DataTypeTok{minsize =} \DecValTok{50}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Across-season distribution of maxNT captured for each sample plot in
this experiment is shown in Fig. 2b. The data has a bimodal distribution
which is reflected in the splitting value (18.7 \(^\circ C\)) for the
Plackett-Luce trees in Fig. 2c. The upper node splits with 49 summer
days (SU). We can interpret these results as that differences in growing
performance of common beans is led by a considerable amount of diurnal
temperature above a warmer threshold of 30 \(^\circ C\) (in this case
\textgreater70\% of the growing days) and warmer nights (\textgreater{}
18.7 \(^\circ C\)).

@@

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-Aguilar2005}{}%
Aguilar, E., Peterson, T. C., Obando, P. R., Frutos, R., Retana, J. A.,
Solera, M., \ldots{} Mayorga, R. (2005). Changes in precipitation and
temperature extremes in Central America and northern South America,
1961--2003. \emph{Journal of Geophysical Research}, \emph{110}(D23),
D23107.
doi:\href{https://doi.org/10.1029/2005JD006119}{10.1029/2005JD006119}

\leavevmode\hypertarget{ref-Brouwer1986}{}%
Brouwer, C., \& Heibloem, M. (1986). \emph{Irrigation water management:
Irrigation water needs} (Training m). Rome, Italy: Food; Agriculture
Organization of The United Nations. Retrieved from
\url{http://www.fao.org/3/S2022E/s2022e00.htm}

\leavevmode\hypertarget{ref-deMedeiros2016}{}%
de Medeiros, G. A., Daniel, L. A., \& Fengler, F. H. (2016). Growth,
Development, and Water Consumption of Irrigated Bean Crop Related to
Growing Degree-Days on Different Soil Tillage Systems in Southeast
Brazil. \emph{International Journal of Agronomy}, \emph{2016}, 8065985.
doi:\href{https://doi.org/10.1155/2016/8065985}{10.1155/2016/8065985}

\leavevmode\hypertarget{ref-deSousa2018}{}%
de Sousa, K., Casanoves, F., Sellare, J., Ospina, A., Suchini, J. G.,
Aguilar, A., \& Mercado, L. (2018). How climate awareness influences
farmers' adaptation decisions in Central America? \emph{Journal of Rural
Studies}, \emph{64}, 11--19.
doi:\href{https://doi.org/10.1016/j.jrurstud.2018.09.018}{10.1016/j.jrurstud.2018.09.018}

\leavevmode\hypertarget{ref-Kehel2016}{}%
Kehel, Z., Crossa, J., \& Reynolds, M. (2016). Identifying Climate
Patterns during the Crop-Growing Cycle from 30 Years of CIMMYT Elite
Spring Wheat International Yield Trials. In A. Bari, A. B. Damania, M.
Mackay, \& S. Dayanandan (Eds.), \emph{Applied mathematics and omics to
assess crop genetic resources for climate change adaptive traits} (pp.
151--174). CRC Press.

\leavevmode\hypertarget{ref-YLiu2018}{}%
Liu, Y., \& El-Kassaby, Y. A. (2018). Evapotranspiration and favorable
growing degree-days are key to tree height growth and ecosystem
functioning: Meta-analyses of Pacific Northwest historical data.
\emph{Scientific Reports}, \emph{8228}.
doi:\href{https://doi.org/10.1038/s41598-018-26681-1}{10.1038/s41598-018-26681-1}

\leavevmode\hypertarget{ref-sf}{}%
Pebesma, E. (2018). Simple Features for R: Standardized Support for
Spatial Vector Data. \emph{The R Journal}, \emph{10}(1), 439--446.
doi:\href{https://doi.org/10.32614/RJ-2018-009}{10.32614/RJ-2018-009}

\leavevmode\hypertarget{ref-Prentice1992}{}%
Prentice, I. C., Cramer, W., Harrison, S. P., Leemans, R., Monserud, R.
A., \& Solomon, A. M. (1992). Special Paper: A Global Biome Model Based
on Plant Physiology and Dominance, Soil Properties and Climate.
\emph{Journal of Biogeography}, \emph{19}(2), 117.
doi:\href{https://doi.org/10.2307/2845499}{10.2307/2845499}

\leavevmode\hypertarget{ref-RCoreTeam}{}%
R Core Team. (2020). R: A language and environment for statistical
computing. version 4.0.0. Vienna, Austria: CRAN R Project. Retrieved
from \url{https://www.r-project.org/}

\leavevmode\hypertarget{ref-PlantEcology}{}%
Schulze, E.-D., Beck, E., Buchmann, N., Clemens, S., Müller-Hohenstein,
K., \& Scherer-Lorenzen, M. (2019). \emph{Plant Ecology} (Second Edi, p.
925). Berlin, Germany: Springer.
doi:\href{https://doi.org/10.1007/978-3-662-56233-8}{10.1007/978-3-662-56233-8}

\leavevmode\hypertarget{ref-Turner2020}{}%
Turner, H. L., van Etten, J., Firth, D., \& Kosmidis, I. (2020).
Modelling rankings in R: the PlackettLuce package. \emph{Computational
Statistics}.
doi:\href{https://doi.org/10.1007/s00180-020-00959-3}{10.1007/s00180-020-00959-3}

\leavevmode\hypertarget{ref-vanEtten2019}{}%
van Etten, J., de Sousa, K., Aguilar, A., Barrios, M., Coto, A.,
Dell'Acqua, M., \ldots{} Steinke, J. (2019). Crop variety management for
climate adaptation supported by citizen science. \emph{Proceedings of
the National Academy of Sciences}, \emph{116}(10), 4194--4199.
doi:\href{https://doi.org/10.1073/pnas.1813720116}{10.1073/pnas.1813720116}

\leavevmode\hypertarget{ref-Wan2015}{}%
Wan, Z., Hook, S., \& Hulley, G. (2015). MYD11A1 MODIS/Aqua Land Surface
Temperature/Emissivity Daily L3 Global 1km SIN Grid V006 {[}Data set{]}.
\emph{NASA EOSDIS LP DAAC}.
doi:\href{https://doi.org/10.5067/MODIS/MYD11A1.006}{10.5067/MODIS/MYD11A1.006}

\leavevmode\hypertarget{ref-JWu2013}{}%
Wu, J., Guan, D., Yuan, F., Wang, A., \& Jin, C. (2013). Soil
Temperature Triggers the Onset of Photosynthesis in Korean Pine.
\emph{PLoS ONE}, \emph{8}(6), e65401.
doi:\href{https://doi.org/10.1371/journal.pone.0065401}{10.1371/journal.pone.0065401}
\end{cslreferences}

@@

\begin{figure}
\includegraphics[width=0.8\linewidth]{cbean} \caption{Fig. 1. Application of climatrends functions to support the analysis of a citizen-science data testing 11 common bean varieties in Nicaragua. (A) Days required to reach 900 growing-degree days from planting date calculated using the function GDD(). (B) Maximum night temperature (°C) distributed across seasons computed using the function temperature(). (C) Plackett-Luce Tree showing the probability of one common bean variety has to win against the others (axys X) in three different nodes splitted with the summer days (day temperature > 30 °C) and maximum night temperature (°C). Note: the first season (primera, Pr) spans from May to August, the second (postrera, Po) from September to October, and the third (apante, Ap) from November to January.}\label{fig:fig_cbean}
\end{figure}
